---
title: "eDNA COImg2 Analysis — Rance Estuary"
output: html_document
date: "`r Sys.Date()`"
---

# ============================================================
# TABLE OF CONTENTS
# ============================================================
#   BLOC 0  — Configuration and Variable Settings
#   BLOC 1  — Setup, import, phyloseq, cleaning, functions
#   BLOC 2  — Alpha per filter (iNEXT) — Arthropoda, Mollusca, Annelida, Cnidaria
#   BLOC 3  — NMDS beta-diversity + envfit (Species matrix + Phylum vectors)
#   BLOC 5  — NMDS per capsule (TorpDNA, Venturi, Waterra) without Lyvet
#   BLOC 6  — Taxonomic composition (RRA + POO) by Phylum and Species
#   BLOC 7  — Beta-diversity partition (betapart) per capsule without Lyvet
#   BLOC 8  — Distance-decay Mantel tests (2 figures)
#   BLOC SUPP — POO comparison + NMDS (Chateaubriand & Pont_Saint_Hubert)
# ============================================================

# ============================================================
# BLOC 0 — Configuration and Variable Settings
# ============================================================

```{r bloc0_configuration, include=FALSE}

# ------------------------------------------------------------
# Paths
# ------------------------------------------------------------
rep_racine <- "C:/Users/efoucaud/Desktop/R_Data_Manipulatio/COImg2/Metabar/Phyloseq_COImg2"

# ------------------------------------------------------------
# Data file names
# ------------------------------------------------------------
fichier_motus   <- "clean_motus_2026.txt"
fichier_reads   <- "clean_reads_2026.txt"
fichier_pcrs    <- "clean_pcrs_2026.txt"
fichier_samples <- "clean_samples_2026.txt"

# ------------------------------------------------------------
# 0.1 — Target phylums
# ------------------------------------------------------------
phylums_cibles <- c("Arthropoda", "Annelida", "Mollusca", "Cnidaria")
pct_identity_threshold <- 97

# ------------------------------------------------------------
# 0.2 — PCRs and samples to exclude
# ------------------------------------------------------------
pcrs_blacklist   <- c("Rance41xR3", "Rance24xR3", "Rance36xR3", "Rance43xR3")
sample_blacklist <- c("Rance002")
samples_um_exclus <- c("Rance036", "Rance037", "Rance038", "Rance039")
sites_exclus_geo  <- c("Vieux_Banc", "Buharats")

# ------------------------------------------------------------
# 0.3 — Site order and labels
# ------------------------------------------------------------
site_order <- c("La_plate", "Bizeux", "Chateaubriand_Richardais",
                "Landriais", "Pont_Saint_Hubert", "Lyvet")

site_labels <- c(
  "La_plate"                 = "La Plate\n(-10 km)",
  "Bizeux"                   = "Bizeux\n(-2 km)",
  "Chateaubriand_Richardais" = "Châteaubriand\n(+2 km)",
  "Landriais"                = "Landriais\n(+4.5 km)",
  "Pont_Saint_Hubert"        = "Pont St-Hubert\n(+10 km)",
  "Lyvet"                    = "Lyvet\n(+19.5 km)"
)

site_labels_court <- c(
  "La_plate"                 = "La Plate",
  "Bizeux"                   = "Bizeux",
  "Chateaubriand_Richardais" = "Châteaubriand",
  "Landriais"                = "Landriais",
  "Pont_Saint_Hubert"        = "Pont St-H.",
  "Lyvet"                    = "Lyvet"
)

site_distances_km <- c(-10, -2, 2, 4.5, 10, 19.5)
meta_sites <- data.frame(Site=site_order, Distance=site_distances_km)

# ------------------------------------------------------------
# 0.4 — Capsules and colours
# ------------------------------------------------------------
capsule_colors <- c(
  "Waterra"  = "#e41a1c",
  "Venturi"  = "#377eb8",
  "TorpeDNA" = "#4daf4a",
  "Sterivex" = "#984ea3")

filter_colors     <- capsule_colors
capsules_principales <- c("Waterra", "Venturi", "TorpeDNA")
filters_principales  <- capsules_principales

# ------------------------------------------------------------
# 0.5 — Phylum colours
# ------------------------------------------------------------
phylum_colors <- c(
  "Arthropoda" = "#e6851f",
  "Annelida"   = "#d62728",
  "Mollusca"   = "#9467bd",
  "Cnidaria"   = "#17becf")

# ------------------------------------------------------------
# 0.6 — Statistical parameters
# ------------------------------------------------------------
n_perm_permanova <- 999
n_perm_envfit    <- 9999
n_perm_mantel    <- 9999
n_perm_indval    <- 9999
seed_global      <- 123

pval_envfit_seuil <- 0.05
n_top_envfit      <- 8

# ------------------------------------------------------------
# 0.7 — Blocs to run
# ------------------------------------------------------------
run_bloc2  <- FALSE
run_bloc3  <- FALSE
run_bloc5  <- FALSE
run_bloc6  <- FALSE
run_bloc7  <- TRUE
run_bloc8  <- FALSE
run_supp   <- TRUE

```

# ============================================================
# BLOC 1 — Setup, import, phyloseq, cleaning, functions
# ============================================================

```{r bloc1_setup, include=FALSE}

library(phyloseq)
library(ggplot2)
library(ggrepel)
library(tibble)
library(dplyr)
library(tidyr)
library(vegan)
library(patchwork)
library(betapart)
library(indicspecies)
library(scales)
library(ggnewscale)
library(ggupset)
library(RColorBrewer)
library(iNEXT)
library(cowplot)

setwd(rep_racine)

# ------------------------------------------------------------
# 1.1 — Import
# ------------------------------------------------------------
motus   <- read.table(fichier_motus,   sep="\t", header=TRUE, row.names=1)
reads   <- read.table(fichier_reads,   sep="\t", header=TRUE, row.names=1)
pcrs    <- read.table(fichier_pcrs,    sep="\t", header=TRUE, row.names=1)
samples <- read.table(fichier_samples, sep="\t", header=TRUE, row.names=1)

samples$Site             <- trimws(samples$Site)
samples$Distance_Aval_km <- as.numeric(samples$Distance_Aval_km)
motus$pct_identity       <- as.numeric(gsub(",", ".", motus$pct_identity))

# ------------------------------------------------------------
# 1.2 — Build phyloseq
# ------------------------------------------------------------
pcrs_full <- pcrs %>%
  tibble::rownames_to_column("pcr_id") %>%
  left_join(samples %>% tibble::rownames_to_column("sample_id"),
            by="sample_id") %>%
  column_to_rownames("pcr_id")

OTU <- otu_table(as.matrix(reads), taxa_are_rows=FALSE)
TAX <- tax_table(as.matrix(motus[, c("Cluster","Phylum","Class","Order",
                                      "Family","Genus","Species","pct_identity")]))
SAM <- sample_data(pcrs_full)
physeq_raw <- phyloseq(OTU, TAX, SAM)

# ------------------------------------------------------------
# 1.3 — Geographic and quality cleaning
# ------------------------------------------------------------
physeq_clean <- subset_samples(physeq_raw, !Site %in% sites_exclus_geo)
physeq_clean <- prune_taxa(taxa_sums(physeq_clean) > 0, physeq_clean)
physeq_clean <- prune_samples(sample_sums(physeq_clean) > 0, physeq_clean)

# Remove blacklisted PCRs
physeq_clean <- prune_samples(
  !sample_names(physeq_clean) %in% pcrs_blacklist, physeq_clean)

# Remove blacklisted sample
pcrs_rance002 <- rownames(pcrs)[pcrs$sample_id %in% sample_blacklist]
physeq_clean  <- prune_samples(
  !sample_names(physeq_clean) %in% pcrs_rance002, physeq_clean)
physeq_clean  <- prune_taxa(taxa_sums(physeq_clean) > 0, physeq_clean)

# MAD z-score outlier detection
physeq_clust_tmp <- tax_glom(physeq_clean, taxrank="Cluster", NArm=TRUE)
physeq_clust_tmp <- prune_taxa(taxa_sums(physeq_clust_tmp) > 0, physeq_clust_tmp)
phy_pa_tmp       <- transform_sample_counts(physeq_clust_tmp,
                      function(x) ifelse(x>0,1,0))
obs_tmp          <- estimate_richness(phy_pa_tmp, measures="Observed")
meta_tmp         <- data.frame(sample_data(physeq_clust_tmp))

alpha_check <- data.frame(
  sample    = rownames(obs_tmp),
  Observed  = obs_tmp$Observed,
  sample_id = meta_tmp$sample_id) %>%
  group_by(sample_id) %>%
  mutate(
    med_obs = median(Observed, na.rm=TRUE),
    mad_obs = mad(Observed, na.rm=TRUE),
    z_score = ifelse(mad_obs > 0, abs(Observed-med_obs)/mad_obs, 0),
    outlier = z_score > 5) %>%
  ungroup()

pcrs_mad_out <- alpha_check %>% filter(outlier) %>% pull(sample)
if(length(pcrs_mad_out) > 0) {
  physeq_clean <- prune_samples(
    !sample_names(physeq_clean) %in% pcrs_mad_out, physeq_clean)
  physeq_clean <- prune_taxa(taxa_sums(physeq_clean) > 0, physeq_clean)
}

# Remove TorpeDNA µm comparison samples
pcrs_um_exclus <- rownames(pcrs)[pcrs$sample_id %in% samples_um_exclus]
physeq_clean   <- prune_samples(
  !sample_names(physeq_clean) %in% pcrs_um_exclus, physeq_clean)
physeq_clean   <- prune_taxa(taxa_sums(physeq_clean) > 0, physeq_clean)

# ------------------------------------------------------------
# 1.4 — Target phylums + identity threshold + remove Sterivex
# ------------------------------------------------------------
physeq_tax <- subset_taxa(physeq_clean, Phylum %in% phylums_cibles)
physeq_tax <- prune_taxa(taxa_sums(physeq_tax) > 0, physeq_tax)

tax_df_tmp <- as.data.frame(tax_table(physeq_tax))
asv_97     <- rownames(tax_df_tmp)[
  as.numeric(tax_df_tmp$pct_identity) >= pct_identity_threshold]
physeq_tax <- prune_taxa(asv_97, physeq_tax)
physeq_tax <- prune_samples(
  sample_data(physeq_tax)$Capsule != "Sterivex", physeq_tax)
physeq_tax <- prune_taxa(taxa_sums(physeq_tax) > 0, physeq_tax)
physeq_tax <- prune_samples(sample_sums(physeq_tax) > 0, physeq_tax)

# Species-level object
physeq_sp <- tax_glom(physeq_tax, taxrank="Species", NArm=TRUE)
physeq_sp <- prune_taxa(taxa_sums(physeq_sp) > 0, physeq_sp)
physeq_sp <- prune_samples(sample_sums(physeq_sp) > 0, physeq_sp)

tax_sp                <- as.data.frame(tax_table(physeq_sp))
tax_sp$Species_full   <- paste(tax_sp$Genus, tax_sp$Species)
taxa_names(physeq_sp) <- make.unique(tax_sp$Species_full)

cat("=== FINAL OBJECTS ===\n")
cat("physeq_tax :", nsamples(physeq_tax), "PCRs |", ntaxa(physeq_tax), "ASVs\n")
cat("physeq_sp  :", nsamples(physeq_sp),  "PCRs |", ntaxa(physeq_sp),  "species\n")
cat("Phylum distribution:\n")
print(table(as.data.frame(tax_table(physeq_sp))$Phylum))

# ------------------------------------------------------------
# 1.5 — Shared functions
# ------------------------------------------------------------

get_nmds_scores <- function(nmds_obj, physeq_obj, niveau_label) {
  scores <- as.data.frame(nmds_obj$points)
  colnames(scores) <- c("NMDS1","NMDS2")
  meta <- data.frame(sample_data(physeq_obj))
  meta$sample   <- rownames(meta)
  scores$sample <- rownames(scores)
  scores %>%
    left_join(meta, by="sample") %>%
    mutate(Site    = factor(trimws(Site), levels=site_order),
           Capsule = factor(Capsule),
           niveau  = niveau_label)
}

run_permanova <- function(dist_obj, physeq_obj, label) {
  meta_p      <- data.frame(sample_data(physeq_obj))
  meta_p$Site <- trimws(meta_p$Site)
  cat("\n========================================\n")
  cat("PERMANOVA —", label, "\n========================================\n")
  print(adonis2(dist_obj ~ Site + Capsule,
                data=meta_p, permutations=n_perm_permanova, by="margin"))
  cat("\nPERMDISP by Site:\n")
  bd <- betadisper(dist_obj, meta_p$Site)
  print(permutest(bd, permutations=n_perm_permanova))
}

run_envfit_env <- function(nmds_obj, physeq_obj) {
  meta <- data.frame(sample_data(physeq_obj))
  nmds_scores <- nmds_obj$points
  common      <- intersect(rownames(nmds_scores), rownames(meta))
  env_df <- data.frame(
    Distance_Aval_km = as.numeric(meta[common, "Distance_Aval_km"]),
    row.names        = common)
  set.seed(seed_global)
  fit_env <- envfit(nmds_scores[common,], env_df,
                    permutations=n_perm_envfit, na.rm=TRUE)
  vec <- as.data.frame(scores(fit_env, display="vectors"))
  vec$taxon <- rownames(vec)
  vec$pval  <- fit_env$vectors$pvals
  vec$r2    <- fit_env$vectors$r
  cat("Distance_Aval_km: R²=", round(vec$r2, 3),
      "| p=", round(vec$pval, 4), "\n")
  return(vec)
}

project_phylum <- function(physeq_pa_obj, nmds_obj,
                             physeq_raw_obj=NULL,
                             pval_threshold=pval_envfit_seuil) {
  otu_mat <- as.data.frame(otu_table(physeq_pa_obj))
  if(taxa_are_rows(physeq_pa_obj)) otu_mat <- t(otu_mat)
  tax_source <- if(!is.null(physeq_raw_obj)) physeq_raw_obj else physeq_pa_obj
  tax_df     <- as.data.frame(tax_table(tax_source))

  phy_mat <- sapply(phylums_cibles, function(p) {
    asv_p <- rownames(tax_df)[!is.na(tax_df$Phylum) & tax_df$Phylum == p]
    asv_p <- intersect(asv_p, colnames(otu_mat))
    if(length(asv_p) == 0) return(rep(0, nrow(otu_mat)))
    if(length(asv_p) == 1) as.numeric(otu_mat[,asv_p] > 0)
    else as.numeric(rowSums(otu_mat[,asv_p, drop=FALSE]) > 0)
  })
  phy_mat <- as.data.frame(phy_mat)
  rownames(phy_mat) <- rownames(otu_mat)

  nmds_scores <- nmds_obj$points
  common      <- intersect(rownames(nmds_scores), rownames(phy_mat))
  if(length(common) < 5) { message("Not enough common samples"); return(NULL) }

  phy_mat_sub <- phy_mat[common, , drop=FALSE]
  phy_mat_sub <- phy_mat_sub[, colSums(phy_mat_sub) > 0, drop=FALSE]
  phy_mat_sub <- phy_mat_sub[, apply(phy_mat_sub, 2, var) > 0, drop=FALSE]
  if(ncol(phy_mat_sub) == 0) { message("No Phylum with variance > 0"); return(NULL) }

  set.seed(seed_global)
  fit <- envfit(nmds_scores[common,], phy_mat_sub,
                permutations=n_perm_envfit, na.rm=TRUE)
  vec <- as.data.frame(scores(fit, display="vectors"))
  vec$taxon <- rownames(vec)
  vec$pval  <- fit$vectors$pvals
  vec$r2    <- fit$vectors$r
  cat("=== Phylum projection ===\n")
  print(vec %>% select(taxon, r2, pval) %>%
    mutate(across(where(is.numeric), round, 3)))
  vec %>% filter(pval <= pval_threshold)
}

prep_betapart_matrix <- function(physeq_obj) {
  phy_tax <- tax_glom(physeq_obj, taxrank="Species", NArm=TRUE)
  phy_tax <- prune_taxa(taxa_sums(phy_tax) > 0, phy_tax)
  phy_pa  <- transform_sample_counts(phy_tax, function(x) ifelse(x>0,1,0))
  otu_mat <- as.data.frame(otu_table(phy_pa))
  if(taxa_are_rows(phy_pa)) otu_mat <- t(otu_mat)
  colnames(otu_mat) <- make.unique(colnames(otu_mat))
  meta_df      <- data.frame(sample_data(phy_pa))
  otu_mat$Site <- trimws(meta_df$Site)
  site_mat <- otu_mat %>%
    group_by(Site) %>%
    summarise(across(where(is.numeric), max), .groups="drop") %>%
    as.data.frame()
  rownames(site_mat) <- site_mat$Site; site_mat$Site <- NULL
  site_mat[intersect(site_order, rownames(site_mat)), ]
}

plot_beta_heatmap <- function(dist_mat, titre, palette_high="#08306b") {
  as.matrix(dist_mat) %>%
    as.data.frame() %>%
    tibble::rownames_to_column("Site1") %>%
    tidyr::pivot_longer(-Site1, names_to="Site2", values_to="value") %>%
    mutate(Site1=factor(trimws(Site1), levels=site_order),
           Site2=factor(trimws(Site2), levels=site_order)) %>%
    ggplot(aes(x=Site2, y=Site1, fill=value)) +
    geom_tile(color="white", linewidth=0.5) +
    geom_text(aes(label=round(value, 2)),
              size=3, color="white", fontface="bold") +
    scale_fill_gradient(low="#f7fbff", high=palette_high,
                        name="Value", limits=c(0,1)) +
    scale_x_discrete(labels=site_labels_court) +
    scale_y_discrete(labels=site_labels_court) +
    theme_bw() +
    labs(title=titre, x=NULL, y=NULL) +
    theme(axis.text.x=element_text(size=8, angle=30, hjust=1),
          axis.text.y=element_text(size=8),
          plot.title=element_text(face="bold", size=11))
}

```

# ============================================================
# BLOC 2 — Alpha per filter + Cumulative richness (iNEXT)
# ============================================================

```{r bloc2_alpha_inext, eval=run_bloc2}

# ---- Helper: observed richness ----
calc_alpha_sp <- function(physeq, niveau_label) {
  phy_pa  <- transform_sample_counts(physeq, function(x) ifelse(x>0,1,0))
  obs     <- estimate_richness(phy_pa, measures="Observed")
  meta    <- data.frame(sample_data(physeq))
  meta$sample <- rownames(meta)
  data.frame(sample=rownames(obs), Observed=obs$Observed,
             niveau=niveau_label) %>%
    left_join(meta, by="sample") %>%
    mutate(Site    = factor(trimws(Site), levels=site_order),
           Capsule = factor(Capsule))
}

# ---- iNEXT helpers ----
build_inext_list_global <- function(physeq_obj, capsule_name) {
  idx <- sample_data(physeq_obj)$Capsule == capsule_name
  if(sum(idx) == 0) return(NULL)
  phy_cap <- prune_samples(idx, physeq_obj)
  phy_cap <- prune_taxa(taxa_sums(phy_cap) > 0, phy_cap)
  if(ntaxa(phy_cap) == 0 || nsamples(phy_cap) < 2) return(NULL)
  otu_mat <- as.data.frame(otu_table(phy_cap))
  if(taxa_are_rows(phy_cap)) otu_mat <- t(otu_mat)
  otu_pa  <- (otu_mat > 0) * 1
  freq    <- colSums(otu_pa); freq <- freq[freq > 0]
  if(length(freq) == 0) return(NULL)
  c(nrow(otu_pa), freq)
}

build_inext_list_site <- function(physeq_obj, capsule_name, site_name) {
  meta_df <- data.frame(sample_data(physeq_obj))
  idx     <- meta_df$Capsule == capsule_name &
             trimws(meta_df$Site) == site_name
  if(sum(idx) == 0) return(NULL)
  phy_cap <- prune_samples(idx, physeq_obj)
  phy_cap <- prune_taxa(taxa_sums(phy_cap) > 0, phy_cap)
  if(ntaxa(phy_cap) == 0 || nsamples(phy_cap) < 1) return(NULL)
  otu_mat <- as.data.frame(otu_table(phy_cap))
  if(taxa_are_rows(phy_cap)) otu_mat <- t(otu_mat)
  otu_pa  <- (otu_mat > 0) * 1
  # For Lyvet (1 PCR): use incidence_raw workaround
  n_units <- nrow(otu_pa)
  if(n_units == 1) {
    freq <- as.numeric(otu_pa[1, ])
    freq <- freq[freq > 0]
    if(length(freq) == 0) return(NULL)
    return(c(n_units, freq))
  }
  freq <- colSums(otu_pa); freq <- freq[freq > 0]
  if(length(freq) == 0) return(NULL)
  c(n_units, freq)
}

# ---- Figure 1: Observed richness by site × capsule ----
alpha_sp <- calc_alpha_sp(physeq_sp, "Species")

alpha_sp_plot <- alpha_sp %>%
  mutate(Site    = factor(trimws(Site), levels=site_order),
         Capsule = factor(Capsule, levels=c("TorpeDNA","Venturi","Waterra")))

p_alpha <- ggplot(alpha_sp_plot, aes(x=Site, y=Observed, fill=Capsule)) +
  geom_boxplot(alpha=0.85, outlier.shape=NA,
               position=position_dodge(width=0.75), width=0.65) +
  geom_jitter(aes(color=Capsule),
              position=position_jitterdodge(jitter.width=0.1, dodge.width=0.75),
              size=2, alpha=0.8) +
  scale_x_discrete(labels=site_labels) +
  scale_fill_manual(values=filter_colors[c("TorpeDNA","Venturi","Waterra")]) +
  scale_color_manual(values=capsule_colors[c("TorpeDNA","Venturi","Waterra")]) +
  theme_bw() +
  labs(title    = "Observed richness by site and filter — Species",
       subtitle = "COImg2 | Arthropoda, Mollusca, Annelida, Cnidaria | pct_identity ≥ 97%",
       x="Site (downstream → upstream)", y="Taxonomic Richness",
       fill="Filter", color="Filter") +
  theme(legend.position="bottom",
        axis.text.x=element_text(size=8),
        plot.title=element_text(face="bold", size=11))

print(p_alpha)
pdf(file.path(rep_racine, "Fig_Bloc2_COImg2_Observed_Richness_by_site_filter.pdf"),
    width=10, height=6)
print(p_alpha)
dev.off()

# Kruskal-Wallis
cat("=== Kruskal-Wallis per filter ===\n")
recap_kw <- do.call(rbind, lapply(filters_principales, function(cap) {
  df_c <- alpha_sp %>% filter(Capsule==cap)
  if(length(unique(df_c$Site)) < 2) return(NULL)
  kw <- kruskal.test(Observed~Site, data=df_c)
  data.frame(Capsule=cap, n_sites=length(unique(df_c$Site)),
             n_PCRs=nrow(df_c), chi2=round(kw$statistic,2),
             p_value=format(kw$p.value, scientific=TRUE, digits=3))
}))
print(recap_kw, row.names=FALSE)

# ---- Figure 2: iNEXT global per capsule ----
set.seed(seed_global)
inext_global_list <- Filter(Negate(is.null),
  setNames(lapply(capsules_principales,
                   function(cap) build_inext_list_global(physeq_sp, cap)),
           capsules_principales))

inext_global_out <- iNEXT(inext_global_list,
                           q=0, datatype="incidence_freq",
                           endpoint=NULL, knots=40,
                           se=TRUE, conf=0.95, nboot=200)

inext_global_df         <- inext_global_out$iNextEst$size_based
inext_global_df$Capsule <- inext_global_df$Assemblage

p_inext_global <- ggplot(inext_global_df,
    aes(x=t, y=qD, color=Capsule, fill=Capsule, linetype=Method)) +
  geom_ribbon(aes(ymin=qD.LCL, ymax=qD.UCL), alpha=0.15, color=NA) +
  geom_line(linewidth=1.1) +
  geom_vline(
    data=inext_global_df %>% group_by(Capsule) %>%
      filter(Method=="Observed") %>% slice_tail(n=1),
    aes(xintercept=t, color=Capsule),
    linetype="dotted", linewidth=0.7, alpha=0.7) +
  scale_color_manual(values=capsule_colors[names(inext_global_list)]) +
  scale_fill_manual(values=capsule_colors[names(inext_global_list)]) +
  scale_linetype_manual(values=c("Rarefaction"="solid",
                                  "Observed"="solid",
                                  "Extrapolation"="dashed")) +
  theme_bw() +
  labs(title    = "Global cumulative richness by filter — Species (iNEXT)",
       subtitle = "COImg2 | Rarefaction/Extrapolation | q=0 | 95% CI | Dotted = observed limit",
       x="Cumulative number of sampling units (PCRs)",
       y="Species richness",
       color="Filter", fill="Filter", linetype="Method") +
  theme(plot.title=element_text(face="bold", size=11),
        legend.position="bottom")

print(p_inext_global)
pdf(file.path(rep_racine, "Fig_Bloc2_COImg2_Global_Cumulative_Richness_iNEXT.pdf"),
    width=9, height=6)
print(p_inext_global)
dev.off()

# ---- Figure 3: iNEXT per site × capsule ----
set.seed(seed_global)
inext_site_plots <- lapply(site_order, function(s) {
  site_list <- Filter(Negate(is.null),
    setNames(lapply(capsules_principales,
                     function(cap) build_inext_list_site(physeq_sp, cap, s)),
             capsules_principales))

  caps_valides <- names(site_list)[sapply(site_list, function(x) x[1] >= 2)]

  if(length(site_list) == 0) {
    cat("Site", s, ": no data — skipped\n"); return(NULL) }

  if(length(caps_valides) == 0) {
    # Single-PCR site (e.g. Lyvet): show observed richness as points
    rich_obs <- do.call(rbind, lapply(names(site_list), function(cap) {
      vec <- site_list[[cap]]
      data.frame(Capsule=cap, n_pcr=vec[1], n_sp=length(vec)-1)
    }))
    return(
      ggplot(rich_obs, aes(x=n_pcr, y=n_sp, color=Capsule)) +
        geom_point(size=4, alpha=0.9) +
        scale_color_manual(values=capsule_colors[rich_obs$Capsule]) +
        scale_x_continuous(limits=c(0,3), breaks=0:3) +
        theme_bw(base_size=9) +
        labs(title=site_labels_court[s],
             subtitle="Single PCR — observed richness only",
             x="Nb PCRs", y="Species richness") +
        theme(plot.title=element_text(face="bold", size=9),
              legend.position="none"))
  }

  tryCatch({
    out <- iNEXT(site_list[caps_valides],
                  q=0, datatype="incidence_freq",
                  endpoint=NULL, knots=30,
                  se=TRUE, conf=0.95, nboot=200)
    df <- out$iNextEst$size_based
    df$Capsule <- df$Assemblage
    ggplot(df, aes(x=t, y=qD, color=Capsule, fill=Capsule, linetype=Method)) +
      geom_ribbon(aes(ymin=qD.LCL, ymax=qD.UCL), alpha=0.15, color=NA) +
      geom_line(linewidth=1.0) +
      scale_color_manual(values=capsule_colors[caps_valides]) +
      scale_fill_manual(values=capsule_colors[caps_valides]) +
      scale_linetype_manual(values=c("Rarefaction"="solid",
                                      "Observed"="solid",
                                      "Extrapolation"="dashed")) +
      theme_bw(base_size=9) +
      labs(title=site_labels_court[s], x="Nb PCRs", y="Species richness") +
      theme(plot.title=element_text(face="bold", size=9),
            legend.position="none")
  }, error=function(e) {
    cat("iNEXT error at site", s, ":", e$message, "\n"); NULL
  })
})
names(inext_site_plots) <- site_order

valid_plots <- Filter(Negate(is.null), inext_site_plots)

p_inext_sites <- wrap_plots(valid_plots, ncol=3) +
  plot_annotation(
    title    = "Cumulative richness by site — Filter comparison (iNEXT)",
    subtitle = "COImg2 | Rarefaction/Extrapolation | q=0 | 95% CI | Dashed = extrapolation",
    theme    = theme(plot.title=element_text(face="bold", size=11),
                     plot.subtitle=element_text(size=9, color="grey40"))) &
  theme(legend.position="none")

leg_src <- ggplot(
  data.frame(Capsule=factor(capsules_principales, levels=capsules_principales),
             x=seq_along(capsules_principales), y=1),
  aes(x=x, y=y, color=Capsule, linetype="Rarefaction")) +
  geom_line(linewidth=1) +
  scale_color_manual(values=capsule_colors[capsules_principales], name="Filter") +
  scale_linetype_manual(values=c("Rarefaction"="solid","Extrapolation"="dashed"),
                         name="Method") +
  theme(legend.position="bottom", legend.direction="horizontal")

final_inext_sites <- plot_grid(
  p_inext_sites, get_legend(leg_src),
  ncol=1, rel_heights=c(0.88, 0.12))

print(final_inext_sites)
pdf(file.path(rep_racine, "Fig_Bloc2_COImg2_Cumulative_Richness_by_site_iNEXT.pdf"),
    width=12, height=8)
print(final_inext_sites)
dev.off()

```

# ============================================================
# BLOC 3 — NMDS beta-diversity + envfit
#          Species-level ordination | Phylum vectors + Distance
# ============================================================

```{r bloc3_nmds_envfit, eval=run_bloc3}

# ---- P/A objects ----
phy_pa_sp <- transform_sample_counts(physeq_sp,
                                      function(x) ifelse(x>0,1,0))
phy_pa_sp <- prune_samples(sample_sums(phy_pa_sp) > 0, phy_pa_sp)

phy_pa_sp_nl <- prune_samples(
  sample_data(phy_pa_sp)$Site != "Lyvet", phy_pa_sp)
phy_pa_sp_nl <- prune_taxa(taxa_sums(phy_pa_sp_nl) > 0, phy_pa_sp_nl)

# ---- NMDS with Lyvet ----
set.seed(seed_global)
dist_jac_sp <- distance(phy_pa_sp, method="jaccard", binary=TRUE)
nmds_sp     <- ordinate(phy_pa_sp, method="NMDS", distance=dist_jac_sp)
cat("Stress NMDS (with Lyvet):", round(nmds_sp$stress, 3), "\n")
scores_sp   <- get_nmds_scores(nmds_sp, phy_pa_sp, "Species")
run_permanova(dist_jac_sp, phy_pa_sp, "Species — with Lyvet")

# ---- NMDS without Lyvet ----
set.seed(seed_global)
dist_jac_sp_nl <- distance(phy_pa_sp_nl, method="jaccard", binary=TRUE)
nmds_sp_nl     <- ordinate(phy_pa_sp_nl, method="NMDS", distance=dist_jac_sp_nl)
cat("Stress NMDS (without Lyvet):", round(nmds_sp_nl$stress, 3), "\n")
scores_sp_nl   <- get_nmds_scores(nmds_sp_nl, phy_pa_sp_nl, "Species")
run_permanova(dist_jac_sp_nl, phy_pa_sp_nl, "Species — without Lyvet")

# ---- envfit ----
cat("\n=== Phylum projection — with Lyvet ===\n")
vec_phy_full <- project_phylum(phy_pa_sp, nmds_sp,
                                physeq_raw_obj=physeq_tax)
cat("\n=== Phylum projection — without Lyvet ===\n")
vec_phy_nl   <- project_phylum(phy_pa_sp_nl, nmds_sp_nl,
                                physeq_raw_obj=physeq_tax)
cat("\n=== Distance_Aval_km — with Lyvet ===\n")
env_full     <- run_envfit_env(nmds_sp,    phy_pa_sp)
cat("\n=== Distance_Aval_km — without Lyvet ===\n")
env_nl       <- run_envfit_env(nmds_sp_nl, phy_pa_sp_nl)

# ---- NMDS plot function ----
build_nmds_plot_coi <- function(scores_df, vectors_phy=NULL,
                                 vector_env=NULL,
                                 titre="", color_var="Site",
                                 scale_phy=0.45, scale_env=0.55) {
  max_range   <- max(abs(c(scores_df$NMDS1, scores_df$NMDS2)))
  color_scale <- if(color_var == "Site") scale_color_brewer(palette="Set2")
                 else scale_color_manual(values=capsule_colors)

  p <- ggplot() +
    geom_point(data=scores_df,
               aes(x=NMDS1, y=NMDS2, color=.data[[color_var]]),
               size=2.2, alpha=0.65) +
    stat_ellipse(data=scores_df,
                 aes(x=NMDS1, y=NMDS2,
                     group=.data[[color_var]],
                     color=.data[[color_var]]),
                 linetype=2, linewidth=0.5) +
    color_scale +
    theme_bw(base_size=9) +
    labs(title=titre, x="NMDS1", y="NMDS2", color=color_var) +
    theme(plot.title=element_text(face="bold", size=10))

  # Phylum vectors
  if(!is.null(vectors_phy) && nrow(vectors_phy) > 0) {
    vec_phy_plot <- vectors_phy %>%
      mutate(x_end    = MDS1 * max_range * scale_phy,
             y_end    = MDS2 * max_range * scale_phy,
             fill_col = phylum_colors[taxon])
    p <- p +
      geom_segment(data=vec_phy_plot,
                   aes(x=0, y=0, xend=x_end, yend=y_end),
                   arrow=arrow(length=unit(0.18,"cm"), type="closed"),
                   color=vec_phy_plot$fill_col, linewidth=0.9) +
      geom_label(data=vec_phy_plot,
                 aes(x=x_end*1.2, y=y_end*1.2,
                     label=taxon),
                 fill=vec_phy_plot$fill_col, color="white",
                 size=2.8, fontface="bold",
                 label.padding=unit(0.18,"lines"))
  }

  # Environmental vector
  if(!is.null(vector_env) && nrow(vector_env) > 0) {
    vec_e <- vector_env %>%
      mutate(x_end = MDS1 * max_range * scale_env,
             y_end = MDS2 * max_range * scale_env)
    p <- p +
      ggnewscale::new_scale_fill() +
      geom_segment(data=vec_e,
                   aes(x=0, y=0, xend=x_end, yend=y_end),
                   arrow=arrow(length=unit(0.25,"cm"), type="closed"),
                   color="#08306b", linewidth=1.3) +
      geom_label(data=vec_e,
                 aes(x=x_end*1.18, y=y_end*1.18,
                     label=paste0("Dist. downstream\np=",round(pval,4))),
                 fill="#08306b", color="white",
                 size=2.5, fontface="bold",
                 label.padding=unit(0.18,"lines"))
  }
  return(p)
}

stress_full <- round(nmds_sp$stress, 3)
stress_nl   <- round(nmds_sp_nl$stress, 3)

p_b3_hg <- build_nmds_plot_coi(
  scores_df=scores_sp, vectors_phy=vec_phy_full, vector_env=env_full,
  titre=paste0("With Lyvet — Phyla + Distance (stress=", stress_full, ")"))
p_b3_hd <- build_nmds_plot_coi(
  scores_df=scores_sp_nl, vectors_phy=vec_phy_nl, vector_env=env_nl,
  titre=paste0("Without Lyvet — Phyla + Distance (stress=", stress_nl, ")"))
p_b3_bg <- build_nmds_plot_coi(
  scores_df=scores_sp, vectors_phy=vec_phy_full, vector_env=env_full,
  titre=paste0("With Lyvet — by capsule (stress=", stress_full, ")"),
  color_var="Capsule")
p_b3_bd <- build_nmds_plot_coi(
  scores_df=scores_sp_nl, vectors_phy=vec_phy_nl, vector_env=env_nl,
  titre=paste0("Without Lyvet — by capsule (stress=", stress_nl, ")"),
  color_var="Capsule")

panel_bloc3 <- (p_b3_hg | p_b3_hd) / (p_b3_bg | p_b3_bd) +
  plot_annotation(
    title    = "Beta-diversity COImg2 — NMDS Jaccard P/A (Species)",
    subtitle = "Phylum vectors (coloured) | Distance_Aval_km (dark blue)",
    theme    = theme(plot.title=element_text(face="bold", size=12),
                     plot.subtitle=element_text(size=9, color="grey40"))) &
  theme(legend.position="bottom")

print(panel_bloc3)
pdf(file.path(rep_racine, "Fig_Bloc3_COImg2_NMDS_panel_2x2.pdf"),
    width=14, height=12)
print(panel_bloc3)
dev.off()

```

# ============================================================
# BLOC 5 — NMDS per capsule (without Lyvet)
#          Phylum vectors + Environmental vector + shared legend
# ============================================================

```{r bloc5_nmds_capsule, eval=run_bloc5}

if(!exists("phy_pa_sp")) {
  phy_pa_sp <- transform_sample_counts(physeq_sp,
                                        function(x) ifelse(x>0,1,0))
  phy_pa_sp <- prune_samples(sample_sums(phy_pa_sp)>0, phy_pa_sp)
}

run_nmds_capsule_nl <- function(physeq_pa_obj, capsule_name) {
  meta_df <- data.frame(sample_data(physeq_pa_obj))
  idx     <- meta_df$Capsule == capsule_name &
             trimws(meta_df$Site) != "Lyvet"
  if(sum(idx) < 4) return(NULL)

  phy_cap <- prune_samples(idx, physeq_pa_obj)
  phy_cap <- prune_taxa(taxa_sums(phy_cap) > 0, phy_cap)
  phy_cap <- prune_samples(sample_sums(phy_cap) > 0, phy_cap)

  n_sites <- length(unique(trimws(as.character(sample_data(phy_cap)$Site))))
  n_pcrs  <- nsamples(phy_cap)
  if(n_pcrs < 5 || n_sites < 2) return(NULL)

  set.seed(seed_global)
  dist_jac <- distance(phy_cap, method="jaccard", binary=TRUE)
  nmds_obj <- tryCatch(ordinate(phy_cap, method="NMDS", distance=dist_jac),
                        error=function(e) NULL)
  if(is.null(nmds_obj)) return(NULL)
  cat("Capsule:", capsule_name, "| Stress:", round(nmds_obj$stress,3), "\n")

  scores_df <- get_nmds_scores(nmds_obj, phy_cap, capsule_name)

  # ── Récupérer les PCRs présentes dans le NMDS ──
  pcrs_nmds <- rownames(nmds_obj$points)

  # ── Construire raw_cap depuis physeq_tax (ASV brut) ──
  raw_cap <- prune_samples(
    sample_names(physeq_tax) %in% pcrs_nmds, physeq_tax)
  raw_cap <- prune_taxa(taxa_sums(raw_cap) > 0, raw_cap)

  # ── Construire matrice Phylum P/A directement ──
  # (évite le problème ASV ID vs species name dans project_phylum)
  vec_phy <- tryCatch({
    tax_raw  <- as.data.frame(tax_table(raw_cap))
    otu_raw  <- as.data.frame(otu_table(raw_cap))
    if(taxa_are_rows(raw_cap)) otu_raw <- t(otu_raw)
    otu_raw  <- otu_raw[intersect(rownames(otu_raw), pcrs_nmds), , drop=FALSE]

    phy_mat <- sapply(phylums_cibles, function(p) {
      asv_p <- rownames(tax_raw)[!is.na(tax_raw$Phylum) &
                                   tax_raw$Phylum == p]
      asv_p <- intersect(asv_p, colnames(otu_raw))
      if(length(asv_p) == 0) return(rep(0, nrow(otu_raw)))
      if(length(asv_p) == 1) as.numeric(otu_raw[, asv_p] > 0)
      else as.numeric(rowSums(otu_raw[, asv_p, drop=FALSE]) > 0)
    })
    phy_mat <- as.data.frame(phy_mat)
    rownames(phy_mat) <- rownames(otu_raw)

    # Garder seulement les phylums avec variance > 0
    phy_mat <- phy_mat[, colSums(phy_mat) > 0, drop=FALSE]
    phy_mat <- phy_mat[, apply(phy_mat, 2, var) > 0, drop=FALSE]
    if(ncol(phy_mat) == 0) return(NULL)

    common <- intersect(pcrs_nmds, rownames(phy_mat))
    if(length(common) < 5) return(NULL)

    set.seed(seed_global)
    fit <- envfit(nmds_obj$points[common,], phy_mat[common,],
                  permutations=n_perm_envfit, na.rm=TRUE)
    vec <- as.data.frame(scores(fit, display="vectors"))
    vec$taxon <- rownames(vec)
    vec$pval  <- fit$vectors$pvals
    vec$r2    <- fit$vectors$r
    cat("  Phylum projection (", capsule_name, "):\n")
    print(vec %>% select(taxon,r2,pval) %>%
            mutate(across(where(is.numeric), round, 3)))
    vec %>% filter(pval <= pval_envfit_seuil)
  }, error=function(e) {
    cat("  Phylum projection failed:", e$message, "\n"); NULL })

  vec_env <- tryCatch(run_envfit_env(nmds_obj, phy_cap),
                       error=function(e) NULL)

  list(scores=scores_df, nmds=nmds_obj, dist=dist_jac,
       physeq=phy_cap, vec_phy=vec_phy, vec_env=vec_env,
       filter=capsule_name, n_sites=n_sites)
}

nmds_cap_nl <- lapply(setNames(capsules_principales, capsules_principales),
                       function(cap) run_nmds_capsule_nl(phy_pa_sp, cap))

# ---- Build NMDS plot per capsule ----
# ---- Axes forcés par capsule ----
axes_par_capsule <- list(
  "Waterra"  = list(xlim=c(-2, 2),    ylim=c(-2, 2)),
  "Venturi"  = list(xlim=c(-2, 2),    ylim=c(-2, 2)),
  "TorpeDNA" = list(xlim=c(-2, 2),    ylim=c(-2, 2))
)

build_nmds_capsule_plot_coi <- function(obj, cap_name) {
  if(is.null(obj)) return(NULL)
  stress_val <- round(obj$nmds$stress, 3)

  # Axes spécifiques à cette capsule
  ax  <- axes_par_capsule[[cap_name]]
  xlim_cap <- ax$xlim
  ylim_cap <- ax$ylim

  # Scale vecteurs proportionnelle à la taille de la fenêtre
  max_range <- max(abs(c(xlim_cap, ylim_cap)))

  p <- ggplot() +
    geom_point(data=obj$scores,
               aes(x=NMDS1, y=NMDS2, color=Site),
               size=3.0, alpha=0.80) +
    stat_ellipse(data=obj$scores,
                 aes(x=NMDS1, y=NMDS2, group=Site, color=Site),
                 linetype=2, linewidth=0.7) +
    scale_color_brewer(palette="Set2", name="Site",
                       labels=site_labels_court[
                         levels(droplevels(obj$scores$Site))]) +
    theme_bw(base_size=10) +
    labs(title    = cap_name,
         subtitle = paste0("Stress = ", stress_val,
                            " | ", obj$n_sites, " sites | Without Lyvet"),
         x="NMDS1", y="NMDS2") +
    coord_fixed(ratio=1,
                xlim=xlim_cap,
                ylim=ylim_cap) +
    theme(plot.title    = element_text(face="bold", size=11),
          plot.subtitle = element_text(size=8, color="grey40"),
          legend.position = "none",
          panel.grid.minor = element_blank())

  # Environmental vector
  if(!is.null(obj$vec_env) && nrow(obj$vec_env) > 0) {
    scale_env <- max_range * 0.80
    vec_e <- obj$vec_env %>%
      mutate(x_end = MDS1 * scale_env,
             y_end = MDS2 * scale_env)
    p <- p +
      geom_segment(data=vec_e,
                   aes(x=0, y=0, xend=x_end, yend=y_end),
                   arrow=arrow(length=unit(0.28,"cm"), type="closed"),
                   color="#08306b", linewidth=1.4) +
      geom_label(data=vec_e,
                 aes(x=x_end * 1.20,
                     y=y_end * 1.20,
                     label=paste0("Dist. downstream\np=",
                                   round(pval, 4))),
                 fill="#08306b", color="white",
                 size=2.8, fontface="bold",
                 label.padding=unit(0.22,"lines"))
  }

  # Phylum vectors
  if(!is.null(obj$vec_phy) && nrow(obj$vec_phy) > 0) {
    scale_phy <- max_range * 0.55
    vec_p <- obj$vec_phy %>%
      mutate(x_end    = MDS1 * scale_phy,
             y_end    = MDS2 * scale_phy,
             fill_col = phylum_colors[taxon])
    p <- p +
      geom_segment(data=vec_p,
                   aes(x=0, y=0, xend=x_end, yend=y_end),
                   arrow=arrow(length=unit(0.20,"cm"), type="closed"),
                   color=vec_p$fill_col, linewidth=1.1) +
      geom_label_repel(data=vec_p,
                       aes(x=x_end, y=y_end, label=taxon),
                       fill=vec_p$fill_col, color="white",
                       size=2.8, fontface="bold",
                       label.padding  = unit(0.20,"lines"),
                       box.padding    = unit(0.80,"lines"),
                       point.padding  = unit(0.40,"lines"),
                       force=15, max.overlaps=Inf,
                       min.segment.length=0,
                       segment.color=vec_p$fill_col,
                       segment.linewidth=0.6,
                       alpha=0.95, seed=42)
  }
  return(p)
}

plots_cap <- lapply(capsules_principales, function(cap)
  build_nmds_capsule_plot_coi(nmds_cap_nl[[cap]], cap))
names(plots_cap) <- capsules_principales

# ---- Shared legend ----
scores_for_legend <- do.call(rbind,
  lapply(nmds_cap_nl, function(obj) if(!is.null(obj)) obj$scores else NULL))
sites_present <- levels(droplevels(
  factor(trimws(scores_for_legend$Site), levels=site_order)))

leg_src5 <- ggplot(
    scores_for_legend %>%
      mutate(Site=factor(trimws(Site), levels=site_order)),
    aes(x=NMDS1, y=NMDS2, color=Site)) +
  geom_point(size=3) +
  scale_color_brewer(palette="Set2", name="Site",
                     labels=site_labels_court[sites_present]) +
  theme(legend.position="bottom", legend.direction="horizontal",
        legend.text=element_text(size=9),
        legend.title=element_text(size=9, face="bold"))

nmds_row5 <- wrap_plots(Filter(Negate(is.null), plots_cap), ncol=3) +
  plot_annotation(
    title    = "NMDS per capsule — Without Lyvet (COImg2 | Jaccard P/A)",
    subtitle = "Phylum vectors (coloured) | Distance_Aval_km (dark blue) | No Lyvet",
    theme    = theme(plot.title=element_text(face="bold", size=12),
                     plot.subtitle=element_text(size=9, color="grey40")))

final_panel5 <- cowplot::plot_grid(
  nmds_row5,
  cowplot::get_legend(leg_src5),
  ncol=1, rel_heights=c(0.88, 0.12))

print(final_panel5)
pdf(file.path(rep_racine, "Fig_Bloc5_COImg2_NMDS_3capsules_without_Lyvet.pdf"),
    width=15, height=7)
print(final_panel5)
dev.off()

# PERMANOVA per capsule
cat("\n=== PERMANOVA per capsule — without Lyvet ===\n")
for(cap in capsules_principales) {
  obj <- nmds_cap_nl[[cap]]; if(is.null(obj)) next
  meta_p <- data.frame(sample_data(obj$physeq))
  meta_p$Site <- trimws(meta_p$Site)
  if(length(unique(meta_p$Site)) < 2) next
  cat("\nCapsule:", cap, "\n")
  tryCatch(print(adonis2(obj$dist ~ Site, data=meta_p,
                          permutations=n_perm_permanova)),
           error=function(e) cat("PERMANOVA not possible\n"))
}

```

# ============================================================
# BLOC 6 — Taxonomic composition (RRA + POO)
#          Phylum global + Species per Phylum
# ============================================================

```{r bloc6_composition, eval=run_bloc6}

library(ggupset)

# ---- Phylum-level RRA ----
tax_df_all   <- as.data.frame(tax_table(physeq_tax))
meta_tax_all <- data.frame(sample_data(physeq_tax))
otu_raw      <- as.data.frame(otu_table(physeq_tax))
if(taxa_are_rows(physeq_tax)) otu_raw <- t(otu_raw)
otu_rra_all  <- as.data.frame(t(apply(otu_raw, 1, function(x) x/sum(x)*100)))
otu_pa_all   <- as.data.frame((otu_raw > 0)*1)

rra_phy_df <- as.data.frame(sapply(phylums_cibles, function(phy) {
  cols <- rownames(tax_df_all)[tax_df_all$Phylum == phy]
  cols <- intersect(cols, colnames(otu_rra_all))
  if(length(cols) == 0) return(rep(0, nrow(otu_rra_all)))
  if(length(cols) == 1) otu_rra_all[, cols]
  else rowSums(otu_rra_all[, cols, drop=FALSE])
}))
rra_phy_df$Site <- trimws(meta_tax_all$Site)

poo_phy_df <- as.data.frame(sapply(phylums_cibles, function(phy) {
  cols <- rownames(tax_df_all)[tax_df_all$Phylum == phy]
  cols <- intersect(cols, colnames(otu_pa_all))
  if(length(cols) == 0) return(rep(0, nrow(otu_pa_all)))
  as.numeric(rowSums(otu_pa_all[, cols, drop=FALSE]) > 0)
}))
poo_phy_df$Site <- trimws(meta_tax_all$Site)

rra_phy_long <- rra_phy_df %>%
  group_by(Site) %>%
  summarise(across(all_of(phylums_cibles), mean), .groups="drop") %>%
  mutate(Site=factor(trimws(Site), levels=site_order)) %>%
  tidyr::pivot_longer(-Site, names_to="Phylum", values_to="RRA") %>%
  mutate(Phylum=factor(Phylum, levels=phylums_cibles))

poo_phy_long <- poo_phy_df %>%
  group_by(Site) %>%
  summarise(across(all_of(phylums_cibles), ~sum(.)/n()*100), .groups="drop") %>%
  mutate(Site=factor(trimws(Site), levels=site_order)) %>%
  tidyr::pivot_longer(-Site, names_to="Phylum", values_to="POO") %>%
  group_by(Site) %>%
  mutate(POO_norm=POO/sum(POO)*100) %>% ungroup() %>%
  mutate(Phylum=factor(Phylum, levels=phylums_cibles))

p_rra_phy <- ggplot(rra_phy_long, aes(x=Site, y=RRA, fill=Phylum)) +
  geom_bar(stat="identity", position="fill", width=0.7) +
  scale_x_discrete(labels=site_labels_court) +
  scale_fill_manual(values=phylum_colors) +
  scale_y_continuous(labels=scales::percent_format()) +
  theme_bw() +
  labs(title="Phylum composition by site — RRA",
       subtitle="COImg2 | Read proportion | 4 target Phyla",
       x=NULL, y="Read proportion", fill="Phylum") +
  theme(axis.text.x=element_text(angle=30, hjust=1, size=9),
        plot.title=element_text(face="bold", size=11))

p_poo_phy <- ggplot(poo_phy_long, aes(x=Site, y=POO_norm, fill=Phylum)) +
  geom_bar(stat="identity", position="stack", width=0.7) +
  scale_x_discrete(labels=site_labels_court) +
  scale_fill_manual(values=phylum_colors) +
  scale_y_continuous(limits=c(0,105), breaks=seq(0,100,25)) +
  theme_bw() +
  labs(title="Phylum composition by site — Normalised POO",
       subtitle="COImg2 | Relative contribution of each Phylum to detections",
       x=NULL, y="Normalised occurrence (%)", fill="Phylum") +
  theme(axis.text.x=element_text(angle=30, hjust=1, size=9),
        plot.title=element_text(face="bold", size=11),
        legend.position="bottom")

panel_phy <- (p_rra_phy | p_poo_phy) +
  plot_annotation(
    title="COImg2 — Phylum composition: RRA vs POO",
    subtitle="4 target Phyla | Same colour palette",
    theme=theme(plot.title=element_text(face="bold", size=12)))
print(panel_phy)
pdf(file.path(rep_racine, "Fig_Bloc6_COImg2_Phylum_RRA_POO.pdf"),
    width=12, height=6)
print(panel_phy)
dev.off()

# ---- Species-level RRA + POO per Phylum ----
for(phy in phylums_cibles) {
  physeq_phy <- subset_taxa(physeq_tax, Phylum==phy)
  physeq_phy <- prune_taxa(taxa_sums(physeq_phy)>0, physeq_phy)
  physeq_phy <- prune_samples(sample_sums(physeq_phy)>0, physeq_phy)
  phy_spe    <- tax_glom(physeq_phy, taxrank="Species", NArm=TRUE)
  phy_spe    <- prune_taxa(taxa_sums(phy_spe)>0, phy_spe)
  phy_spe    <- prune_samples(sample_sums(phy_spe)>0, phy_spe)
  n_spe      <- ntaxa(phy_spe)
  if(n_spe == 0) next

  tax_spe    <- as.data.frame(tax_table(phy_spe))
  meta_spe   <- data.frame(sample_data(phy_spe))
  spe_names  <- make.unique(tax_spe$Species)
  top_spe_n  <- min(15, n_spe)

  # RRA
  phy_rra_spe <- transform_sample_counts(phy_spe, function(x) x/sum(x)*100)
  otu_rra_spe <- as.data.frame(otu_table(phy_rra_spe))
  if(taxa_are_rows(phy_rra_spe)) otu_rra_spe <- t(otu_rra_spe)
  colnames(otu_rra_spe) <- spe_names
  otu_rra_spe$Site      <- trimws(meta_spe$Site)

  rra_long <- otu_rra_spe %>%
    group_by(Site) %>%
    summarise(across(all_of(spe_names), mean), .groups="drop") %>%
    mutate(Site=factor(trimws(Site), levels=site_order)) %>%
    tidyr::pivot_longer(-Site, names_to="Species", values_to="RRA")

  top_spes <- rra_long %>%
    group_by(Species) %>% summarise(tot=sum(RRA), .groups="drop") %>%
    arrange(desc(tot)) %>% slice_head(n=top_spe_n) %>% pull(Species)

  rra_plot <- rra_long %>%
    mutate(Species=ifelse(Species %in% top_spes, Species, "Others")) %>%
    group_by(Site, Species) %>% summarise(RRA=sum(RRA), .groups="drop") %>%
    group_by(Site) %>% mutate(RRA_norm=RRA/sum(RRA)*100) %>% ungroup()

  spes_u <- unique(rra_plot$Species[rra_plot$Species != "Others"])
  n_pal  <- length(spes_u)
  pal_spe <- setNames(
    c(colorRampPalette(brewer.pal(min(12,max(3,n_pal)),"Paired"))(n_pal),
      "grey80"),
    c(spes_u, "Others"))

  # POO
  phy_pa_spe_sub <- transform_sample_counts(phy_spe, function(x) ifelse(x>0,1,0))
  otu_poo_spe    <- as.data.frame(otu_table(phy_pa_spe_sub))
  if(taxa_are_rows(phy_pa_spe_sub)) otu_poo_spe <- t(otu_poo_spe)
  colnames(otu_poo_spe) <- spe_names
  otu_poo_spe$Site      <- trimws(data.frame(sample_data(phy_pa_spe_sub))$Site)

  poo_plot <- otu_poo_spe %>%
    group_by(Site) %>%
    summarise(across(all_of(spe_names), ~sum(.)/n()*100), .groups="drop") %>%
    mutate(Site=factor(trimws(Site), levels=site_order)) %>%
    tidyr::pivot_longer(-Site, names_to="Species", values_to="POO") %>%
    mutate(Species=ifelse(Species %in% top_spes, Species, "Others")) %>%
    group_by(Site, Species) %>% summarise(POO=sum(POO), .groups="drop") %>%
    group_by(Site) %>% mutate(POO_norm=POO/sum(POO)*100) %>% ungroup()

  p_rra_spe <- ggplot(rra_plot, aes(x=Site, y=RRA_norm, fill=Species)) +
    geom_bar(stat="identity", position="stack", width=0.7) +
    scale_x_discrete(labels=site_labels_court) +
    scale_fill_manual(values=pal_spe) +
    scale_y_continuous(limits=c(0,105), breaks=seq(0,100,25)) +
    theme_bw() +
    labs(title=paste0(phy, " — Species (RRA)"),
         subtitle=paste0("Top ", top_spe_n, " species | Normalised RRA"),
         x=NULL, y="Normalised RRA (%)", fill="Species") +
    theme(axis.text.x=element_text(angle=30, hjust=1, size=8),
          legend.text=element_text(size=6.5, face="italic"),
          legend.key.size=unit(0.35,"cm"),
          plot.title=element_text(face="bold", size=10))

  p_poo_spe <- ggplot(poo_plot, aes(x=Site, y=POO_norm, fill=Species)) +
    geom_bar(stat="identity", position="stack", width=0.7) +
    scale_x_discrete(labels=site_labels_court) +
    scale_fill_manual(values=pal_spe) +
    scale_y_continuous(limits=c(0,105), breaks=seq(0,100,25)) +
    theme_bw() +
    labs(title=paste0(phy, " — Species (POO)"),
         subtitle="Normalised occurrence frequency | Same palette as RRA",
         x=NULL, y="Normalised occurrence (%)", fill="Species") +
    theme(axis.text.x=element_text(angle=30, hjust=1, size=8),
          legend.text=element_text(size=6.5, face="italic"),
          legend.key.size=unit(0.35,"cm"),
          plot.title=element_text(face="bold", size=10))

  panel_spe <- (p_rra_spe | p_poo_spe) +
    plot_annotation(
      title=paste0("COImg2 — ", phy, " — Species composition: RRA vs POO"),
      subtitle="Same colour palette | Same top species",
      theme=theme(plot.title=element_text(face="bold", size=12),
                  plot.subtitle=element_text(size=9, color="grey40")))
  print(panel_spe)
  pdf(file.path(rep_racine,
                paste0("Fig_Bloc6_COImg2_Species_RRA_POO_", phy, ".pdf")),
      width=13, height=6)
  print(panel_spe)
  dev.off()
}

```

# ============================================================
# BLOC 7 — Beta-diversity partition (betapart) per capsule
#          WITHOUT Lyvet
# ============================================================

```{r bloc7_betapart, eval=run_bloc7}

library(betapart)

# Global without Lyvet
physeq_tax_nl <- prune_samples(
  sample_data(physeq_tax)$Site != "Lyvet", physeq_tax)
physeq_tax_nl <- prune_taxa(taxa_sums(physeq_tax_nl)>0, physeq_tax_nl)

mat_sp_nl <- prep_betapart_matrix(physeq_tax_nl)
beta_nl   <- beta.multi(mat_sp_nl, index.family="sorensen")

cat("\n=== Global beta-diversity partition — WITHOUT Lyvet ===\n")
data.frame(
  Bsor     = round(beta_nl$beta.SOR, 3),
  Bsim     = round(beta_nl$beta.SIM, 3),
  Bnes     = round(beta_nl$beta.SNE, 3),
  Pct_turn = round(beta_nl$beta.SIM/beta_nl$beta.SOR*100, 1)) %>% print()

beta_pair_nl <- beta.pair(mat_sp_nl, index.family="sorensen")

# Wilcoxon global sans Lyvet
bsim_nl_v    <- as.vector(as.dist(beta_pair_nl$beta.sim))
bsne_nl_v    <- as.vector(as.dist(beta_pair_nl$beta.sne))
tw_global_nl <- wilcox.test(bsim_nl_v, bsne_nl_v,
                             paired=TRUE, alternative="greater")
cat("\n=== Wilcoxon Turnover > Nestedness — Global — WITHOUT Lyvet ===\n")
cat("p =", round(tw_global_nl$p.value, 4), "\n")

p_b1 <- plot_beta_heatmap(beta_pair_nl$beta.sor,
  "Total dissimilarity (Bsor) — without Lyvet", "#252525")
p_b2 <- plot_beta_heatmap(beta_pair_nl$beta.sim,
  "Turnover (Bsim) — without Lyvet", "#7f0000")
p_b3 <- plot_beta_heatmap(beta_pair_nl$beta.sne,
  "Nestedness (Bnes) — without Lyvet", "#00441b")

panel_beta_global <- (p_b1 | p_b2 | p_b3) +
  plot_annotation(
    title    = "COImg2 — Beta-diversity partition — Without Lyvet",
    subtitle = "Bsor = Bsim + Bnes | Sørensen | Species level",
    theme    = theme(plot.title=element_text(face="bold", size=12)))
print(panel_beta_global)
pdf(file.path(rep_racine, "Fig_Bloc7_COImg2_Betapart_Global_without_Lyvet.pdf"),
    width=14, height=5)
print(panel_beta_global)
dev.off()

# ---- Per capsule WITHOUT Lyvet ----
cat("\n=== Betapart per capsule — WITHOUT Lyvet ===\n")

recap_bp_cap <- do.call(rbind, lapply(capsules_principales, function(cap) {
  phy_cap <- prune_samples(
    sample_data(physeq_tax_nl)$Capsule == cap, physeq_tax_nl)
  phy_cap <- prune_taxa(taxa_sums(phy_cap)>0, phy_cap)
  phy_cap <- prune_samples(sample_sums(phy_cap)>0, phy_cap)
  if(nsamples(phy_cap) < 4) return(NULL)

  mat <- tryCatch(prep_betapart_matrix(phy_cap), error=function(e) NULL)
  if(is.null(mat) || nrow(mat) < 3) return(NULL)

  bm <- tryCatch(beta.multi(mat, index.family="sorensen"), error=function(e) NULL)
  if(is.null(bm)) return(NULL)

  bp <- tryCatch(beta.pair(mat, index.family="sorensen"), error=function(e) NULL)

  # Wilcoxon turnover vs nestedness
  if(!is.null(bp)) {
    tw <- tryCatch(
      wilcox.test(as.vector(as.dist(bp$beta.sim)),
                  as.vector(as.dist(bp$beta.sne)),
                  paired=TRUE, alternative="greater"),
      error=function(e) NULL)
    p_wilcox <- if(!is.null(tw)) round(tw$p.value, 4) else NA
  } else { p_wilcox <- NA }

  cat("\nCapsule:", cap, "\n")
  cat("  Bsor=", round(bm$beta.SOR,3), "Bsim=", round(bm$beta.SIM,3),
      "Bnes=", round(bm$beta.SNE,3),
      "| %Turn=", round(bm$beta.SIM/bm$beta.SOR*100,1),
      "| Wilcoxon p=", p_wilcox, "\n")

  # Heatmaps per capsule
  if(!is.null(bp)) {
    ph_sim <- plot_beta_heatmap(bp$beta.sim,
      paste0("Turnover (Bsim) — ", cap, " — without Lyvet"), "#7f0000")
    ph_sne <- plot_beta_heatmap(bp$beta.sne,
      paste0("Nestedness (Bnes) — ", cap, " — without Lyvet"), "#00441b")
    panel_cap <- (ph_sim | ph_sne) +
      plot_annotation(
        title    = paste0("COImg2 — Beta-diversity partition — ", cap, " — Without Lyvet"),
        subtitle = paste0("Bsor=", round(bm$beta.SOR,3),
                           " | Bsim=", round(bm$beta.SIM,3),
                           " | Bnes=", round(bm$beta.SNE,3),
                           " | %Turnover=", round(bm$beta.SIM/bm$beta.SOR*100,1), "%"),
        theme=theme(plot.title=element_text(face="bold", size=11),
                    plot.subtitle=element_text(size=9, color="grey40")))
    print(panel_cap)
    pdf(file.path(rep_racine,
                  paste0("Fig_Bloc7_COImg2_Betapart_", cap, "_without_Lyvet.pdf")),
        width=10, height=5)
    print(panel_cap)
    dev.off()

    # --- PDF séparés : Turnover seul ---
    pdf(file.path(rep_racine,
                  paste0("Fig_Bloc7_COImg2_Betapart_", cap, "_Turnover_without_Lyvet.pdf")),
        width=6, height=5)
    print(ph_sim)
    dev.off()

    # --- PDF séparés : Nestedness seul ---
    pdf(file.path(rep_racine,
                  paste0("Fig_Bloc7_COImg2_Betapart_", cap, "_Nestedness_without_Lyvet.pdf")),
        width=6, height=5)
    print(ph_sne)
    dev.off()

    cat("  --> PDFs Turnover et Nestedness exportés séparément pour", cap, "\n")
  }

  data.frame(Capsule=cap, n_sites=nrow(mat),
             Bsor=round(bm$beta.SOR,3), Bsim=round(bm$beta.SIM,3),
             Bnes=round(bm$beta.SNE,3),
             Pct_turn=round(bm$beta.SIM/bm$beta.SOR*100,1),
             Wilcoxon_p=p_wilcox)
}))

cat("\n=== Betapart summary per capsule — without Lyvet ===\n")
print(recap_bp_cap, row.names=FALSE)

```

# ============================================================
# BLOC 8 — Distance-decay — 2 figures in 1 PDF
# Figure 1: Global WITH Lyvet
# Figure 2: Global + per capsule WITHOUT Lyvet (with 95% CI)
# ============================================================

```{r bloc8_mantel, eval=run_bloc8}

library(indicspecies)

if(!exists("mat_sp_nl")) {
  physeq_tax_nl <- prune_samples(
    sample_data(physeq_tax)$Site != "Lyvet", physeq_tax)
  physeq_tax_nl <- prune_taxa(taxa_sums(physeq_tax_nl)>0, physeq_tax_nl)
  mat_sp_nl <- prep_betapart_matrix(physeq_tax_nl)
}
mat_sp <- prep_betapart_matrix(physeq_tax)

run_mantel <- function(mat_pa, label) {
  sites_p  <- intersect(site_order, rownames(mat_pa))
  dist_geo <- as.dist(outer(
    meta_sites$Distance[match(sites_p, meta_sites$Site)],
    meta_sites$Distance[match(sites_p, meta_sites$Site)],
    function(a,b) abs(a-b)))
  dist_eco <- vegdist(mat_pa[sites_p,], method="bray", binary=TRUE)
  cat("\n=== Mantel —", label, "===\n")
  set.seed(seed_global)
  mtest <- mantel(dist_eco, dist_geo, method="pearson",
                  permutations=n_perm_mantel)
  print(mtest)
  data.frame(label=label, r_mantel=round(mtest$statistic,3),
             p_value=round(mtest$signif,4))
}

mantel_full <- run_mantel(mat_sp,    "Global — with Lyvet")
mantel_nl   <- run_mantel(mat_sp_nl, "Global — without Lyvet")

# ---- Figure 1: Global WITH Lyvet ----
sites_full  <- rownames(mat_sp)
dist_sp_geo <- outer(meta_sites$Distance, meta_sites$Distance,
                      function(a,b) abs(a-b))
dist_sp_eco <- as.matrix(vegdist(mat_sp, method="bray", binary=TRUE))

pairs_full <- data.frame(
  Site1    = sites_full[combn(length(sites_full),2)[1,]],
  Site2    = sites_full[combn(length(sites_full),2)[2,]],
  dist_geo = dist_sp_geo[lower.tri(dist_sp_geo)],
  dist_eco = dist_sp_eco[lower.tri(dist_sp_eco)]) %>%
  mutate(Type=ifelse(Site1=="Lyvet"|Site2=="Lyvet",
                      "Involves Lyvet","Without Lyvet"))

p_fig1_mantel <- ggplot(pairs_full, aes(x=dist_geo, y=dist_eco, color=Type)) +
  geom_point(size=2.5, alpha=0.8) +
  geom_smooth(data=pairs_full, aes(x=dist_geo, y=dist_eco),
              method="lm", se=TRUE, color="black", linewidth=1,
              inherit.aes=FALSE) +
  scale_color_manual(values=c("Involves Lyvet"="#e41a1c","Without Lyvet"="grey50")) +
  scale_y_continuous(limits=c(0,1.05)) +
  theme_bw(base_size=11) +
  labs(title    = "Distance-decay — Global with Lyvet",
       subtitle = paste0("Mantel r=", mantel_full$r_mantel,
                          " | p=", mantel_full$p_value,
                          " | Sørensen P/A | Species level | COImg2"),
       x="Geographic distance (km)", y="Dissimilarity (Sørensen)",
       color="Pair type") +
  theme(plot.title=element_text(face="bold", size=11),
        legend.position="bottom")

# ---- Figure 2: Global + per capsule WITHOUT Lyvet (IC) ----
build_mat_cap <- function(physeq_obj, capsule_name) {
  phy_pa  <- transform_sample_counts(physeq_obj, function(x) ifelse(x>0,1,0))
  otu_mat <- as.data.frame(otu_table(phy_pa))
  if(taxa_are_rows(phy_pa)) otu_mat <- t(otu_mat)
  meta_dd         <- data.frame(sample_data(phy_pa))
  otu_mat$Site    <- trimws(meta_dd$Site)
  otu_mat$Capsule <- meta_dd$Capsule
  mat <- otu_mat %>%
    filter(Capsule==capsule_name, Site!="Lyvet") %>%
    select(-Capsule) %>%
    group_by(Site) %>%
    summarise(across(where(is.numeric), max), .groups="drop") %>%
    as.data.frame()
  rownames(mat) <- mat$Site; mat$Site <- NULL
  sites_ok <- intersect(site_order, rownames(mat))
  mat <- mat[sites_ok, colSums(mat[sites_ok,])>0, drop=FALSE]
  if(nrow(mat) < 3) return(NULL)
  mat
}

get_pairs_cap <- function(mat_cap, capsule_name) {
  sites_c  <- rownames(mat_cap)
  dists    <- meta_sites$Distance[match(sites_c, meta_sites$Site)]
  dist_geo <- as.dist(outer(dists, dists, function(a,b) abs(a-b)))
  dist_eco <- vegdist(mat_cap, method="bray", binary=TRUE)
  set.seed(seed_global)
  mt <- mantel(dist_eco, dist_geo, method="pearson",
               permutations=n_perm_mantel)
  cat("\nMantel", capsule_name, "without Lyvet: r=",
      round(mt$statistic,3), "p=", round(mt$signif,4), "\n")
  geo_mat <- as.matrix(dist_geo)
  eco_mat <- as.matrix(dist_eco)
  data.frame(dist_geo=geo_mat[lower.tri(geo_mat)],
             dist_eco=eco_mat[lower.tri(eco_mat)],
             Capsule=capsule_name,
             r_mantel=round(mt$statistic,3),
             p_mantel=round(mt$signif,4))
}

# Global pairs without Lyvet
sites_nl    <- rownames(mat_sp_nl)
dist_geo_nl <- outer(
  meta_sites$Distance[match(sites_nl, meta_sites$Site)],
  meta_sites$Distance[match(sites_nl, meta_sites$Site)],
  function(a,b) abs(a-b))
dist_eco_nl <- as.matrix(vegdist(mat_sp_nl, method="bray", binary=TRUE))

pairs_global_nl <- data.frame(
  dist_geo = dist_geo_nl[lower.tri(dist_geo_nl)],
  dist_eco = dist_eco_nl[lower.tri(dist_eco_nl)],
  Capsule  = "Global",
  r_mantel = mantel_nl$r_mantel,
  p_mantel = mantel_nl$p_value)

# Per capsule pairs
all_cap_pairs <- do.call(rbind, lapply(capsules_principales, function(cap) {
  mat_cap <- build_mat_cap(physeq_sp, cap)
  if(is.null(mat_cap)) return(NULL)
  tryCatch(get_pairs_cap(mat_cap, cap), error=function(e) NULL)
}))

pairs_fig2 <- rbind(pairs_global_nl, all_cap_pairs) %>%
  mutate(Capsule=factor(Capsule, levels=c("Global", capsules_principales)))

colors_fig2 <- c("Global"="black",
                  "Waterra"="#e41a1c",
                  "Venturi"="#377eb8",
                  "TorpeDNA"="#4daf4a")

mantel_labels <- pairs_fig2 %>%
  distinct(Capsule, r_mantel, p_mantel) %>%
  mutate(lab=paste0(Capsule, " r=", r_mantel, " p=", p_mantel)) %>%
  pull(lab) %>% paste(collapse=" | ")

p_fig2_mantel <- ggplot(pairs_fig2,
    aes(x=dist_geo, y=dist_eco, color=Capsule, fill=Capsule)) +
  geom_point(size=2.0, alpha=0.55) +
  geom_smooth(method="lm", se=TRUE, linewidth=1.0, alpha=0.12) +
  scale_color_manual(values=colors_fig2, name="Method") +
  scale_fill_manual(values=colors_fig2,  name="Method") +
  scale_y_continuous(limits=c(0,1.05)) +
  theme_bw(base_size=11) +
  labs(title    = "Distance-decay — Without Lyvet: Global + per capsule",
       subtitle = mantel_labels,
       x="Geographic distance (km)", y="Dissimilarity (Sørensen)") +
  theme(plot.title=element_text(face="bold", size=11),
        plot.subtitle=element_text(size=7.5, color="grey40"),
        legend.position="bottom")

# ---- Assemble 2 figures in 1 PDF ----
panel_mantel <- p_fig1_mantel | p_fig2_mantel
print(panel_mantel +
  plot_annotation(
    title    = "Distance-decay — Mantel tests (COImg2 | Sørensen P/A | Species level)",
    subtitle = "Left: global with Lyvet | Right: without Lyvet — Global + per capsule with 95% CI",
    theme    = theme(plot.title=element_text(face="bold", size=13),
                     plot.subtitle=element_text(size=9, color="grey40"),
                     plot.background=element_rect(fill="white", color=NA))))

pdf(file.path(rep_racine, "Fig_Bloc8_COImg2_Distance_decay_2panels.pdf"),
    width=14, height=6)
print(panel_mantel +
  plot_annotation(
    title    = "Distance-decay — Mantel tests (COImg2 | Sørensen P/A | Species level)",
    subtitle = "Left: global with Lyvet | Right: without Lyvet — Global + per capsule with 95% CI",
    theme    = theme(plot.title=element_text(face="bold", size=13),
                     plot.subtitle=element_text(size=9, color="grey40"),
                     plot.background=element_rect(fill="white", color=NA))))
dev.off()

# ---- IndVal per site ----
phy_sp_iv  <- tax_glom(physeq_tax, taxrank="Species", NArm=TRUE)
phy_sp_iv  <- prune_taxa(taxa_sums(phy_sp_iv)>0, phy_sp_iv)
phy_pa_iv  <- transform_sample_counts(phy_sp_iv, function(x) ifelse(x>0,1,0))
otu_iv     <- as.data.frame(otu_table(phy_pa_iv))
if(taxa_are_rows(phy_pa_iv)) otu_iv <- t(otu_iv)
tax_iv     <- as.data.frame(tax_table(phy_pa_iv))
sp_full_iv <- make.unique(tax_iv$Species)
colnames(otu_iv) <- sp_full_iv

meta_iv      <- data.frame(sample_data(phy_pa_iv))
meta_iv$Site <- trimws(meta_iv$Site)
site_factor  <- factor(meta_iv$Site, levels=site_order)

set.seed(seed_global)
cat("\n=== IndVal per site ===\n")
indval_site <- multipatt(otu_iv, site_factor,
                          func="IndVal.g", control=how(nperm=n_perm_indval))
summary(indval_site, alpha=0.05, indvalcomp=TRUE)

s_cols   <- grep("^s\\.", colnames(indval_site$sign), value=TRUE)
site_nms <- trimws(gsub("^s\\.", "", s_cols))

indval_df <- data.frame(Species=rownames(indval_site$sign),
                         indval_site$sign, stringsAsFactors=FALSE) %>%
  filter(!is.na(p.value), p.value<=0.05) %>%
  mutate(n_sites    = rowSums(across(all_of(s_cols))),
         Site_assoc = apply(across(all_of(s_cols)), 1, function(x) {
           idx <- which(x==1)
           if(length(idx)==0) return(NA)
           paste(site_nms[idx], collapse=" + ")})) %>%
  filter(n_sites==1) %>%
  mutate(Site_assoc=factor(trimws(Site_assoc), levels=site_order),
         stat=round(stat,3), p.value=round(p.value,4),
         p_label=case_when(
           p.value<0.0001 ~ "p=1e-04",
           p.value<0.001  ~ paste0("p=",formatC(p.value, format="e", digits=0)),
           TRUE           ~ paste0("p=",p.value))) %>%
  arrange(Site_assoc, desc(stat))

cat("\nNb indicator species:", nrow(indval_df), "\n")
if(nrow(indval_df) > 0) {
  site_colors_iv <- setNames(
    brewer.pal(max(3, length(site_order)),"Set2"), site_order)
  p_indval <- ggplot(indval_df,
      aes(x=stat, y=reorder(Species, stat), fill=Site_assoc)) +
    geom_bar(stat="identity", width=0.7) +
    geom_text(aes(label=p_label), hjust=-0.1, size=2.8, color="grey30") +
    facet_wrap(~Site_assoc, scales="free_y", ncol=2,
               labeller=as_labeller(site_labels_court)) +
    scale_fill_manual(values=site_colors_iv) +
    theme_bw() +
    labs(title    = "Indicator species per site (IndVal.g) — COImg2",
         subtitle = "Species associated with a single site | p ≤ 0.05",
         x="IndVal", y=NULL) +
    theme(legend.position="none",
          strip.text=element_text(face="bold", size=9),
          plot.title=element_text(face="bold", size=11))
  print(p_indval)
  pdf(file.path(rep_racine, "Fig_Bloc8_COImg2_IndVal_per_site.pdf"),
      width=10, height=8)
  print(p_indval)
  dev.off()
}

```

# ============================================================
# BLOC SUPPLEMENTAIRE — POO comparison + NMDS
# Sites: Chateaubriand_Richardais & Pont_Saint_Hubert
# COImg2 | Order level for POO | Capsule-level NMDS
# ============================================================

```{r bloc_supp_coi, eval=run_supp}

library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(RColorBrewer)
library(cowplot)
library(ggrepel)

# ============================================================
# PARAMETRES
# ============================================================

sites_focus  <- c("Chateaubriand_Richardais", "Pont_Saint_Hubert")
labels_focus <- c(
  "Chateaubriand_Richardais" = "Châteaubriand",
  "Pont_Saint_Hubert"        = "Pont St-Hubert")
n_top_poo  <- 10
col_autres <- "grey70"

library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(RColorBrewer)
library(cowplot)

# ============================================================
# PARAMETRES
# ============================================================

sites_focus  <- c("Chateaubriand_Richardais", "Pont_Saint_Hubert")
labels_focus <- c(
  "Chateaubriand_Richardais" = "Châteaubriand",
  "Pont_Saint_Hubert"        = "Pont St-Hubert")
n_top_poo  <- 10
col_autres <- "grey70"

# ============================================================
# CALCUL POO AU NIVEAU ESPECE PUIS AGREGATION PAR ORDER
# ============================================================

# ---- Sous-ensemble 2 sites depuis physeq_tax (ASV brut avec Phylum) ----
phy_focus_tax <- prune_samples(
  trimws(sample_data(physeq_tax)$Site) %in% sites_focus,
  physeq_tax)
phy_focus_tax <- prune_taxa(taxa_sums(phy_focus_tax) > 0, phy_focus_tax)

# Filtrer les ASVs sans Order, Species ou Phylum valides
phy_focus_tax <- subset_taxa(phy_focus_tax,
  !is.na(Order)   & Order   != "" &
  !is.na(Species) & Species != "" &
  !is.na(Phylum)  & Phylum  != "")

cat("Nb ASVs après filtrage NA:", ntaxa(phy_focus_tax), "\n")

# ---- Agglomération au niveau Espèce ----
phy_focus_sp <- tax_glom(phy_focus_tax, taxrank="Species", NArm=TRUE)
phy_focus_sp <- prune_taxa(taxa_sums(phy_focus_sp) > 0, phy_focus_sp)
phy_focus_sp <- prune_samples(sample_sums(phy_focus_sp) > 0, phy_focus_sp)

cat("Nb espèces après tax_glom:", ntaxa(phy_focus_sp), "\n")

# ---- Table taxonomique espèces avec Order et Phylum ----
tax_sp <- as.data.frame(tax_table(phy_focus_sp))
tax_sp$asv_id      <- rownames(tax_sp)
tax_sp$Order_label <- paste0(tax_sp$Order, " (", tax_sp$Phylum, ")")

cat("Nb Orders uniques:", length(unique(tax_sp$Order_label)), "\n")

# ---- Matrice P/A espèce ----
otu_sp <- as.data.frame(otu_table(phy_focus_sp))
if(taxa_are_rows(phy_focus_sp)) otu_sp <- t(otu_sp)
# lignes = PCRs, colonnes = asv_id

meta_sp         <- data.frame(sample_data(phy_focus_sp))
meta_sp$pcr_id  <- rownames(meta_sp)

otu_pa_sp <- as.data.frame((otu_sp > 0) * 1)

# ---- Calcul POO espèce par site × capsule ----
# POO espèce = nb PCRs où l'espèce est détectée / nb PCRs total du groupe × 100
otu_pa_sp$Site    <- trimws(meta_sp[rownames(otu_pa_sp), "Site"])
otu_pa_sp$Capsule <- meta_sp[rownames(otu_pa_sp), "Capsule"]

poo_sp_long <- otu_pa_sp %>%
  filter(Capsule %in% capsules_principales,
         Site    %in% sites_focus) %>%
  group_by(Site, Capsule) %>%
  summarise(
    n_pcr = n(),
    across(where(is.numeric), ~ sum(.) / n() * 100),
    .groups="drop") %>%
  select(-n_pcr) %>%
  tidyr::pivot_longer(
    cols     = where(is.numeric),
    names_to = "asv_id",
    values_to = "POO_sp") %>%
  mutate(
    Site    = factor(trimws(Site), levels=sites_focus),
    Capsule = factor(Capsule, levels=capsules_principales))

# ---- Joindre l'Order_label à chaque espèce ----
poo_sp_long <- poo_sp_long %>%
  left_join(tax_sp %>% select(asv_id, Order_label), by="asv_id")

# ---- Agréger POO par Order (somme des POO des espèces) ----
# Cette somme reflète bien la "contribution" de chaque Order :
# un Order avec beaucoup d'espèces fréquentes aura une POO élevée
poo_ord_long <- poo_sp_long %>%
  filter(!is.na(Order_label)) %>%
  group_by(Site, Capsule, Order_label) %>%
  summarise(POO_order = sum(POO_sp), .groups="drop")

cat("\n=== POO range par Order ===\n")
cat("Min POO:", round(min(poo_ord_long$POO_order),2), "\n")
cat("Max POO:", round(max(poo_ord_long$POO_order),2), "\n")

# ---- Top 10 Orders par capsule (POO moyenne sur les 2 sites) ----
top_ord_par_capsule <- poo_ord_long %>%
  group_by(Capsule, Order_label) %>%
  summarise(POO_mean=mean(POO_order), .groups="drop") %>%
  group_by(Capsule) %>%
  slice_max(POO_mean, n=n_top_poo, with_ties=FALSE) %>%
  ungroup()

cat("\n=== Top", n_top_poo, "orders per capsule ===\n")
print(top_ord_par_capsule %>%
        arrange(Capsule, desc(POO_mean)) %>%
        mutate(POO_mean=round(POO_mean,1)))

# ---- Classement top10 / Others ----
poo_plot_df <- poo_ord_long %>%
  left_join(
    top_ord_par_capsule %>% select(Capsule, Order_label) %>%
      mutate(is_top=TRUE),
    by=c("Capsule","Order_label")) %>%
  mutate(
    is_top     = replace_na(is_top, FALSE),
    Order_disp = ifelse(is_top, Order_label, "Others")) %>%
  group_by(Site, Capsule, Order_disp) %>%
  summarise(POO_order=sum(POO_order), .groups="drop")

# ---- Grille complète ----
all_ord_labs <- unique(poo_plot_df$Order_disp)
grid_complet <- expand.grid(
  Site       = factor(sites_focus, levels=sites_focus),
  Capsule    = factor(capsules_principales, levels=capsules_principales),
  Order_disp = all_ord_labs,
  stringsAsFactors=FALSE)

poo_plot_df <- grid_complet %>%
  left_join(poo_plot_df, by=c("Site","Capsule","Order_disp")) %>%
  mutate(POO_order=replace_na(POO_order, 0))

# ---- Normalisation → somme = 100% par site × capsule ----
poo_plot_df <- poo_plot_df %>%
  group_by(Site, Capsule) %>%
  mutate(POO_norm=POO_order/sum(POO_order)*100) %>%
  ungroup()

# ---- Supprimer Others si 0% ----
poo_plot_df_filtered <- poo_plot_df %>%
  filter(!(Order_disp=="Others" & POO_norm==0))

# ---- Nb espèces par site × capsule (pour annotation) ----
n_sp_detected <- poo_sp_long %>%
  filter(POO_sp > 0) %>%
  group_by(Site, Capsule) %>%
  summarise(n_sp_det=n_distinct(asv_id), .groups="drop") %>%
  mutate(
    Site    = factor(trimws(Site), levels=sites_focus),
    Capsule = factor(Capsule, levels=capsules_principales))

cat("\n=== Nb species detected per site × capsule ===\n")
print(n_sp_detected)

# ---- Palette ----
all_top_ord <- sort(unique(
  poo_plot_df$Order_disp[poo_plot_df$Order_disp != "Others"]))
n_ord_pal   <- length(all_top_ord)

cat("\nNb unique orders in legend:", n_ord_pal, "\n")
cat("Orders:\n"); print(all_top_ord)

pal_orders <- c(
  setNames(
    colorRampPalette(brewer.pal(12,"Paired"))(n_ord_pal),
    all_top_ord),
  "Others" = col_autres)

ord_levels <- c("Others", all_top_ord)

poo_plot_df_filtered <- poo_plot_df_filtered %>%
  mutate(Order_disp=factor(Order_disp, levels=ord_levels))

# ============================================================
# GRAPHIQUE
# ============================================================

make_poo_panel_coi <- function(cap_name) {
  df_cap  <- poo_plot_df_filtered %>% filter(Capsule==cap_name)

  # Annotation : nb espèces détectées sous chaque barre
  annot_n <- n_sp_detected %>%
    filter(Capsule==cap_name) %>%
    mutate(label=paste0("n=", n_sp_det, " sp."))

  ggplot(df_cap, aes(x=Site, y=POO_norm, fill=Order_disp)) +
    geom_col(position="stack", width=0.6,
             color="white", linewidth=0.4) +
    geom_text(
      data        = annot_n,
      mapping     = aes(x=Site, y=-4, label=label, fill=NULL),
      size        = 3.0, color="grey30", fontface="italic",
      inherit.aes = FALSE) +
    scale_x_discrete(labels=labels_focus) +
    scale_fill_manual(values=pal_orders, name="Order (Phylum)") +
    scale_y_continuous(
      limits = c(-8, 105),
      breaks = seq(0, 100, 25),
      labels = function(x) paste0(x, "%"),
      expand = expansion(mult=c(0,0))) +
    theme_bw(base_size=10) +
    labs(title=cap_name, x=NULL, y="Normalised POO (%)") +
    theme(
      plot.title         = element_text(face="bold", size=11, hjust=0.5),
      axis.text.x        = element_text(size=9, face="bold",
                                         color=filter_colors[cap_name]),
      axis.text.y        = element_text(size=8),
      legend.position    = "none",
      panel.grid.major.x = element_blank(),
      plot.margin        = margin(6, 6, 12, 6))
}

p_wat <- make_poo_panel_coi("Waterra")
p_ven <- make_poo_panel_coi("Venturi")
p_tor <- make_poo_panel_coi("TorpeDNA")

leg_poo <- get_legend(
  ggplot(poo_plot_df_filtered,
         aes(x=Site, y=POO_norm, fill=Order_disp)) +
    geom_col() +
    scale_fill_manual(
      values = pal_orders,
      name   = "Order (Phylum) — Top 10 per capsule",
      breaks = ord_levels) +
    guides(fill=guide_legend(
      ncol=3, title.position="top", title.hjust=0.5,
      byrow=TRUE, override.aes=list(size=5, color=NA))) +
    theme(
      legend.position   = "bottom",
      legend.background = element_rect(fill="white", color="grey85",
                                        linewidth=0.4),
      legend.margin     = margin(6, 8, 6, 8),
      legend.text       = element_text(size=8.5),
      legend.title      = element_text(size=10, face="bold"),
      legend.key.size   = unit(0.55,"cm"),
      legend.key        = element_rect(color="grey70", linewidth=0.3)))

panel_poo <- plot_grid(p_wat, p_ven, p_tor,
                        nrow=1, align="h", axis="tb")

fig_poo <- plot_grid(
  ggdraw() + draw_label(
    "Normalised POO by capsule — Châteaubriand vs Pont St-Hubert (COImg2 | Order level)",
    fontface="bold", size=13, hjust=0.5) +
    theme(plot.margin=margin(4,0,0,0)),
  ggdraw() + draw_label(
    paste0("Top ", n_top_poo,
           " orders by cumulative species POO | Format: Order (Phylum) | n = species detected"),
    size=9, color="grey40", hjust=0.5),
  panel_poo,
  leg_poo,
  ncol=1,
  rel_heights=c(0.05, 0.04, 0.60, 0.31))

print(fig_poo)

pdf(file.path(rep_racine,
              "Fig_Supp_COImg2_POO_Chateaubriand_Pont_top10_Order.pdf"),
    width=14, height=9)
print(fig_poo)
dev.off()

cat("\n=== END — POO Order level bloc ===\n")

# ============================================================
# PARTIE 1b — PERMANOVA par site (Châteaubriand & Pont St-Hubert)
# Variable explicative : Capsule | Distance : Jaccard P/A
# ============================================================

cat("\n=== PERMANOVA par site — Capsule effect (COImg2) ===\n")

sites_supp <- c("Chateaubriand_Richardais", "Pont_Saint_Hubert")

if(!exists("phy_pa_sp")) {
  phy_pa_sp <- transform_sample_counts(physeq_sp,
                                        function(x) ifelse(x > 0, 1, 0))
  phy_pa_sp <- prune_samples(sample_sums(phy_pa_sp) > 0, phy_pa_sp)
}

permanova_supp_results <- lapply(sites_supp, function(site_name) {
  phy_site <- prune_samples(
    trimws(sample_data(phy_pa_sp)$Site) == site_name &
    sample_data(phy_pa_sp)$Capsule %in% capsules_principales,
    phy_pa_sp)
  phy_site <- prune_taxa(taxa_sums(phy_site) > 0, phy_site)
  phy_site <- prune_samples(sample_sums(phy_site) > 0, phy_site)

  n_pcr  <- nsamples(phy_site)
  n_caps <- length(unique(sample_data(phy_site)$Capsule))
  cat("\nSite:", site_name, "| PCRs:", n_pcr, "| Capsules:", n_caps, "\n")

  if(n_pcr < 4 || n_caps < 2) {
    cat("  --> Not enough samples for PERMANOVA\n")
    return(NULL)
  }

  set.seed(seed_global)
  dist_jac <- distance(phy_site, method="jaccard", binary=TRUE)
  meta_p   <- data.frame(sample_data(phy_site))
  meta_p$Capsule <- factor(meta_p$Capsule, levels=capsules_principales)

  cat("\n--- PERMANOVA adonis2 ~ Capsule ---\n")
  perm_res <- adonis2(dist_jac ~ Capsule,
                      data=meta_p,
                      permutations=n_perm_permanova,
                      by="margin")
  print(perm_res)

  cat("\n--- PERMDISP by Capsule ---\n")
  bd <- betadisper(dist_jac, meta_p$Capsule)
  print(permutest(bd, permutations=n_perm_permanova))

  list(site=site_name, dist=dist_jac, meta=meta_p, result=perm_res)
})
names(permanova_supp_results) <- sites_supp

cat("\n=== END — PERMANOVA par site ===\n")

# ============================================================
# PARTIE 2 — NMDS PAR SITE : 2 colonnes
# Gauche  : Châteaubriand | Droite : Pont St-Hubert
# Couleur = Capsule | Légende en bas
# ============================================================

# ---- P/A object ----
if(!exists("phy_pa_sp")) {
  phy_pa_sp <- transform_sample_counts(physeq_sp,
                                        function(x) ifelse(x > 0, 1, 0))
  phy_pa_sp <- prune_samples(sample_sums(phy_pa_sp) > 0, phy_pa_sp)
}

# ---- Fonction NMDS par site ----
build_nmds_site_coi <- function(physeq_pa_obj, site_name, site_label) {
  phy_site <- prune_samples(
    trimws(sample_data(physeq_pa_obj)$Site) == site_name &
    sample_data(physeq_pa_obj)$Capsule %in% capsules_principales,
    physeq_pa_obj)
  phy_site <- prune_taxa(taxa_sums(phy_site) > 0, phy_site)
  phy_site <- prune_samples(sample_sums(phy_site) > 0, phy_site)

  n_pcr  <- nsamples(phy_site)
  n_caps <- length(unique(sample_data(phy_site)$Capsule))
  cat("\nNMDS site:", site_name,
      "| PCRs:", n_pcr, "| Capsules:", n_caps, "\n")

  if(n_pcr < 4 || n_caps < 2) {
    cat("  --> Not enough samples\n")
    return(NULL)
  }

  set.seed(seed_global)
  dist_jac <- distance(phy_site, method="jaccard", binary=TRUE)
  nmds_obj <- tryCatch(
    ordinate(phy_site, method="NMDS", distance=dist_jac),
    error=function(e) { cat("  --> NMDS failed:", e$message, "\n"); NULL })

  if(is.null(nmds_obj)) return(NULL)
  cat("  --> Stress:", round(nmds_obj$stress, 3), "\n")

  scores_df <- as.data.frame(nmds_obj$points)
  colnames(scores_df) <- c("NMDS1", "NMDS2")
  meta_s           <- data.frame(sample_data(phy_site))
  scores_df$sample <- rownames(scores_df)
  meta_s$sample    <- rownames(meta_s)
  scores_df <- scores_df %>%
    left_join(meta_s, by="sample") %>%
    mutate(Capsule = factor(Capsule, levels=capsules_principales))

  list(scores   = scores_df,
       nmds     = nmds_obj,
       stress   = round(nmds_obj$stress, 3),
       site     = site_name,
       label    = site_label,
       n_pcr    = n_pcr)
}

nmds_chat <- build_nmds_site_coi(phy_pa_sp,
                                   "Chateaubriand_Richardais",
                                   "Châteaubriand")
nmds_pont <- build_nmds_site_coi(phy_pa_sp,
                                   "Pont_Saint_Hubert",
                                   "Pont St-Hubert")

# ---- Fonction de tracé NMDS ----
plot_nmds_site_coi <- function(obj) {
  if(is.null(obj)) {
    return(ggplot() +
             annotate("text", x=0.5, y=0.5,
                      label="Not enough data", size=5, color="grey50") +
             theme_void())
  }

  max_range <- max(abs(c(obj$scores$NMDS1, obj$scores$NMDS2))) * 1.3

  ggplot(obj$scores,
         aes(x=NMDS1, y=NMDS2, color=Capsule, shape=Capsule)) +
    stat_ellipse(aes(group=Capsule, color=Capsule),
                 linetype=2, linewidth=0.6,
                 level=0.95, type="t") +
    geom_point(size=3.5, alpha=0.85) +
    scale_color_manual(
      values = capsule_colors[capsules_principales],
      name   = "Capsule") +
    scale_shape_manual(
      values = c("Waterra"=16, "Venturi"=17, "TorpeDNA"=15),
      name   = "Capsule") +
    coord_cartesian(xlim=c(-max_range, max_range),
                    ylim=c(-max_range, max_range)) +
    theme_bw(base_size=11) +
    labs(
      title    = obj$label,
      subtitle = paste0("Stress = ", obj$stress,
                         " | Jaccard P/A | ", obj$n_pcr, " PCRs"),
      x="NMDS1", y="NMDS2") +
    theme(
      plot.title       = element_text(face="bold", size=12, hjust=0.5),
      plot.subtitle    = element_text(size=8, color="grey40", hjust=0.5),
      legend.position  = "none",
      panel.grid.minor = element_blank())
}

p_nmds_chat <- plot_nmds_site_coi(nmds_chat)
p_nmds_pont <- plot_nmds_site_coi(nmds_pont)

# ---- Légende partagée capsules ----
leg_nmds <- get_legend(
  ggplot(
    data.frame(
      Capsule = factor(capsules_principales, levels=capsules_principales),
      x = 1:3, y = 1:3),
    aes(x=x, y=y, color=Capsule, shape=Capsule)) +
    geom_point(size=4) +
    scale_color_manual(
      values = capsule_colors[capsules_principales],
      name   = "Capsule") +
    scale_shape_manual(
      values = c("Waterra"=16, "Venturi"=17, "TorpeDNA"=15),
      name   = "Capsule") +
    guides(color=guide_legend(
      title.position = "top",
      direction      = "horizontal",
      nrow           = 1,
      override.aes   = list(size=4))) +
    theme(
      legend.position = "bottom",
      legend.text     = element_text(size=10),
      legend.title    = element_text(size=10, face="bold")))

# ---- Assemblage : 2 NMDS en colonnes + légende en bas ----
nmds_row <- plot_grid(
  p_nmds_chat, p_nmds_pont,
  nrow=1, align="h", axis="tb",
  labels=c("A","B"), label_size=12)

fig_nmds <- plot_grid(
  ggdraw() + draw_label(
    "NMDS per site — Capsule comparison (COImg2 | Jaccard P/A | Species level)",
    fontface="bold", size=13, hjust=0.5) +
    theme(plot.margin=margin(4, 0, 0, 0)),
  ggdraw() + draw_label(
    "Châteaubriand (+2 km) vs Pont St-Hubert (+10 km) | 90% ellipses per capsule",
    size=9, color="grey40", hjust=0.5),
  nmds_row,
  leg_nmds,
  ncol=1,
  rel_heights=c(0.06, 0.04, 0.78, 0.12))

print(fig_nmds)

pdf(file.path(rep_racine,
              "Fig_Supp_COImg2_NMDS_Chateaubriand_Pont_par_capsule.pdf"),
    width=12, height=7)
print(fig_nmds)
dev.off()

cat("\n=== END — COImg2 supplementary bloc ===\n")

```


